Maximum Entropy Spectral Analysis: an application to gravitational waves data analysis

### 2024-07-01 paper/mesa.tex
Added bibliographic reference

Hence, as an application of MESA, and its implementation in \texttt{memspectrum}, we analyse GW150914\cite{} using a Bayesian framework that allows for the marginalisation of the order  $k$ of the AR($k$) process representing the residuals data stream. Although the inference is essentially unchanged compared to the standard case, see \cite{lalinference}, there are some substantial modification to the likelihood construction. Since MESA is applicable to time-domain data, all calculations prior to the Fourier transform must be performed in time domain, thus increasing the computational cost by a non-negligible amount. We shall refer the time-of-arrival parameter $t_c$ of the GW to the geocenter. At each iteration of the inference algorithm, we sample a vector $\theta\equiv\theta_{GW}\cup k$\footnote{We indicate the set of all GW parameters (component masses, spins, luminosity distance, etc.) with $\theta_{GW}$.}. For each inteferometer $j$, therefore, we need to compute a time-delay $\Delta t_j$ to compute the antenna response functions $F_{j,+}(t+\Delta t_j),F_{j,\times}(t+\Delta t_j)$ as well as the correct time-shift for the GW template 
\begin{align}
h_j(t) =\sum_{p = +,\times}F_{j,p}(t+\Delta t_j)h_{p}(t+\Delta t_j;\theta_{GW})
\end{align}
that we use to compute the time-domain residuals $r_j(t) = d_j(t) - h_j(t)$. We apply \texttt{memspectrum} to $r_j(t)$ with the \emph{fixed} value of $k$ and calculate the detector likelihood for $\tilde{r}_j$ using Eq.~\eqref{eq:Whittle} and PSD as in Eq.~\eqref{eq:MESApsd}. The coherent likelihood is then given by the product of the individual likelihoods. As our analysis template, we adopt the fast machine learning based MLGW model \cite{MLGW}, an aligned spin model trained on TEOBResumS \cite{teobresums}, that has been shown to perform well on LVK events detected during O1 and O2 \cite{MLGW}. Our sampler is a nested sampling algorithm \cite{john_veitch_2020_4109277} and the specific inference model is implemented as part of \texttt{granite}, a dedicated inference model for ground-based interferometric detectors. We compare our results with the combined posterior samples available from GWOSC \cite{GWOSC} and available at \linebreak
\texttt{https://zenodo.org/records/6513631}.

>>>>

Hence, as an application of MESA, and its implementation in \texttt{memspectrum}, we analyse GW150914\cite{gw150914} using a Bayesian framework that allows for the marginalisation of the order  $k$ of the AR($k$) process representing the residuals data stream. Although the inference is essentially unchanged compared to the standard case, see \cite{lalinference}, there are some substantial modification to the likelihood construction. Since MESA is applicable to time-domain data, all calculations prior to the Fourier transform must be performed in time domain, thus increasing the computational cost by a non-negligible amount. We shall refer the time-of-arrival parameter $t_c$ of the GW to the geocenter. At each iteration of the inference algorithm, we sample a vector $\theta\equiv\theta_{GW}\cup k$\footnote{We indicate the set of all GW parameters (component masses, spins, luminosity distance, etc.) with $\theta_{GW}$.}. For each inteferometer $j$, therefore, we need to compute a time-delay $\Delta t_j$ to compute the antenna response functions $F_{j,+}(t+\Delta t_j),F_{j,\times}(t+\Delta t_j)$ as well as the correct time-shift for the GW template 
\begin{align}
h_j(t) =\sum_{p = +,\times}F_{j,p}(t+\Delta t_j)h_{p}(t+\Delta t_j;\theta_{GW})
\end{align}
that we use to compute the time-domain residuals $r_j(t) = d_j(t) - h_j(t)$. We apply \texttt{memspectrum} to $r_j(t)$ with the \emph{fixed} value of $k$ and calculate the detector likelihood for $\tilde{r}_j$ using Eq.~\eqref{eq:Whittle} and PSD as in Eq.~\eqref{eq:MESApsd}. The coherent likelihood is then given by the product of the individual likelihoods. As our analysis template, we adopt the fast machine learning based MLGW model \cite{MLGW}, an aligned spin model trained on TEOBResumS \cite{teobresums}, that has been shown to perform well on LVK events detected during O1 and O2 \cite{MLGW}. Our sampler is a nested sampling algorithm \cite{john_veitch_2020_4109277} and the specific inference model is implemented as part of \texttt{granite}, a dedicated inference model for ground-based interferometric detectors. We compare our results with the combined posterior samples available from GWOSC \cite{GWOSC} and available at \linebreak
\texttt{https://zenodo.org/records/6513631}.
